---
title: "**Survival Analysis of Chargeoff Data**"
subtitle: |
  Samuel Nankervis, Evelyn Zhuang, Kay Ferrin\
  Southern Utah University\
  Fundamentals of Data Science\
  Dr. Rick Brown\
  [DATE]
format: 
  html:
    embed-resources: true # Keep this so the html file is easily shareable.
urlcolor: blue
execute: # Will set these as the default for all code chunks
  eval: true
  echo: true
editor_options: 
  chunk_output_type: console
---
<!--
the organization of the header might need to be updated to look more professional, but I just copied his formatting on the homeworks. also this is how to comment in the quarto document. We could even include a fancy front page like in this link: https://github.com/nmfs-opensci/quarto_titlepages/blob/main/example_1.pdf
-->

# Introduction

<!--
I usually do the introduction last, so I'll leave this blank for now but it should say [something something knowing when loans default is useful so we wanted to know how to do that etc.]
-->
::: {style="text-indent: 30px;"}
This indents a paragraph
:::

# The Data

::: {style="text-indent: 30px;"}
Our dataset came from State Bank of Southern Utah's Loan charge off data. It includes all loans that charged off from 1979 to 2026. While a mixed data set that had both loans that charged off and loans that didn't would be ideal for using cox regression, State Bank of Southern Utah currently doesn't have a dataset with this combined data. Therefore, interpretations of the data will be assuming that the loans issued will charge off at some point, even through this is an unrealistic assumption to make for most real world applications. Despite this limitation of the data, much of the methodology used later in this report can easily be transferred to another data set that does include a mix of data. 
:::

Our original data contains several variables:

**Application Code** ***(categorical)***: The application codes indicate the type of loan on a broad perspective, **Commercial Loans (CC)**, **Installment or traditional Consumer Loans (CI)**, or **Real Estate (CR)**

**Product** ***(categorical)***: the type of loan issued. 

**Interest Rate** ***(continuous)***: the effective annual rate associated with the loan.

**Original Note Date** ***(Discrete)***: The date the loan was issued.

**Original Maturity Date** ***(Discrete)***: The date the loan was intended to mature.

**Current Maturity Date** ***(Discrete)***: The maturity date after loan modifications to assist payment.

**Charge Off Date** ***(Discrete)***: The date the loan was charged off.

**Original Loan Amount** ***(continuous)***: The amount borrowed.

**Total Chargeoff** ***(continuous)***: The amount charged off, or deemed noncollectable at the Charge Off Date.

**LTD Recovery Amount** ***(continuous)***: the amount that was collected after the loan was charged off up until the data was retrieved. 

::: {style="text-indent: 30px;"}
Most of the variables will be useful for predicting survival time, but not in their current form. We can derive the time in days to maturity by subtracting the Charge Off Date by the Original Note Date. Similarly, we can get the original term of the loan by subtracting the Original Maturity Date by the Original Note Date. We can also get an effective monthly rate from the effective annual rate, which will be helpful for time value of money calculations. We will also calculate what the payment per month would be assuming constant interest rate and payments by using the present value of an annuity formula and solving for the payments. These variables will be useful for our data analysis.
:::

Our predictor variables will be:

**Effective Annual Rate (EAR)** ***(continuous)***: the EAR associated with the loan.

**Monthly Payment** ***(continuous)***: the Payments per month of the loan assuming constant payments and interest.
<!--
Note that there may be multicolineariy issues with this specific variable since it is just the result of an equation using the interest rate, term of the loan, and loan amount.
-->

**Time to Maturity** ***(discrete)***: Time in days until the loan was originally intended to mature.

**Original Loan Amount** ***(continuous)***: The amount borrowed.

**Product** ***(categorical)***: the type of loan issued.

**Application Code** ***(categorical)***: The type of loan on a broad perspective, **Commercial Loans (CC)**, **Installment or traditional Consumer Loans (CI)**, or **Real Estate (CR)**

Our response variable will be:

**Time to Charge Off** ***(discrete)***: Time in days until the loan was Charged Off or censored.

::: {style="text-indent: 30px;"}
Now that we have our variables, we need to clean up the data so it doesn't include inaccurate or biased observations. Two types of loan under the Product variable, "Chargeoff Surecash/Visa" and "Surecash", do not have accurate Maturity Dates, so they will be removed from the dataset. Another Product, "Interest Only Home Equity", often has no Maturity Date. This is because Interest Only Home Equity loans do not have a principal reduction These will be removed from the dataset to avoid outliers. The same goes for "Home Equity 97" Products since they also have no end date. I also removed loans with zero interest rate because these were most likely null values. 

<!--
_NOTE-S: I actually don't know how best to handle these. right now these loans are set to mature in the year 9999, but I don't know if having massive outlines like that will hurt our model performance so I think removing them will be best.
_NOTE-K: Agreed!
-->

We also need to remove any loans that had an  Original Maturity Date after when we retrieved the data on March 5th, 2026. We will also remove any loans with Original Note Dates after 2018. The reason we are excluding loans that were set to end after when we collected the data is because we don't have data on all the loans, but only the ones that have already been charged off. This causes a problem for us since we don't have data on loans that will be charged off in the future. Because of this, the data we do have on recent loans will be biased towards loans that charge off quickly after being issued. We also have the issue that there is no data on loans yet to be charged off, so we have no censored data, a requirement of our Cox regression model. So we will simulate censored data by marking all loans that have not charged off by 2018 as censored. This will give us some data that has a piece of the survival time, but not the exact date that the loan will be charged off.

<!--
_NOTE-K: by that, do you mean you kept ones that were not charged off yet in 2015 as the "censored" data? ie still "alive" at time of data collection/no further data on when chargeoff occurred? Or do you mean you removed them completely?

_NOTE-S: Sorry I probably could have made that a bit more clear, I've re written it to hopefully be better. I didn't actually remove those dates I just added a variable to loans that were charged off after 2015 that marked them as censored. We don't need to remove the date that these variables charged off, although we wouldn't know when exactly they would have been charged off had we collected data at that point and known that they would charge off at some point. Hopefully that clarifies things. 
-->

To reiterate, we would not need to transform the data this way if we had data on all loans. With our data cleaned, we will now need to find a statistical model to predict survival time from our predictor variables. 
:::

# Cox Regression
<!--
This section will talk about what cox regression is and how it works. I think it should transition well with the last section.


Here's some general info on Cox regression I'm putting in here so I don't loose the link:

https://numiqo.com/tutorial/cox-regression
https://diogoribeiro7.github.io/finance/data%20science/risk%20modeling/survival_analysis_applied_finance/#case-studies
https://www.kaggle.com/datasets/mannekuntanagendra/survival-analysis-in-r-studio
https://www.youtube.com/watch?v=vtOj8-3bI3M

_NOTE-K: Here's more:
https://www.sciencedirect.com/science/article/pii/S1756231716300639 
-->

<!--
_NOTE-K: I added a bit about age to this section.
-->
  The time it takes for a loan to chargeoff is a type of "survival time," as it is the time until an event that cannot happen multiple times and ends an interval (of loan collection). Survival time probabilities can be successfully used for risk modeling in similar cases, such as for creating expected credit loss profiles (https://diogoribeiro7.github.io/finance/data%20science/risk%20modeling/survival_analysis_applied_finance/#case-studies). Survival time analysis can be done through Cox Regression, also called Cox Proportional Hazards Survival Regression. This method allows us to predict the changing rates associated with the hazard (loan chargeoff in our case) over time, alongside relevant covariates and groupings like we might apply to any other regression. 
  In order to perform a Cox regression, we need to know an amount of time, in our case the time from start of loan until chargeoff, which is used as our response variable. It is also important, with Cox regression specifically, to acknowledge censored data. Censored data is data for which the outcome of interest (loan chargeoff, death, etc) has not yet occurred. This data is not useless, as we know up to a certain point that they "survived" (in our case, that they have not been charged off), but we do not know how long it would take them to get to that point. Cox regression can use this type of data alongside data where the final outcome (loan chargeoff) has already occurred, unlike some other forms of regression. We need a column in our dataset marking particular observations with whether or not the event occurred, so that we do not mistakenly represent censored data as uncensored (the time interval being representative of time until event instead of representing how long we know the event did not yet occur). 
  Another special feature of the Cox proportional hazards regression is the fact that age can be used as a predictor. Age is complicated to use alongside time series regressions because it changes along with time automatically, but the Cox regression is useful in that it is set up to include this. For the loan charge-off data, age is represented by the amount of time from loan initiation to loan charge-off (or until 2015, which is our censor date), the effective age of the loan.
  We can use both categorical and quantitative predictor variables in Cox regression, with time automatically acting as a predictor as well. The probability of default at the beginning is not the same as after some time has passed, and the Cox regression format accounts for this.
  
Mathematics of Cox Regression
  The following definitions of terms and distributions are relevant for understanding survival curves in general:
  Time before event of interest: $T \geq 0$ , in our case time to loan chargeoff.
  Survival function: $S (t) = P(T > t), 0 < t < \infty$ This function represents the probability that an individual survives longer than a particular amount of time. For our data, that means the probability a loan has not been charged off after a certain amount of time.
   Probability density function: $f(t) = -\dfrac{dS(t)}{dt}$ is the negative derivative of the survival probability function with respect to time. It represents the frequency of the event of interest per period of time.
   Hazard function: $h(t) = \lim_{\delta t \to 0^+} \dfrac{P(t\leq T < t + \delta t|T\geq t)}{\delta t}$, also represented as $h(t) = f(t)/S(t)$, represents the risk of the event of interest occurring instantly given it has not happened as of time $t$. In other words, what is the risk of the event occurring now if it hasn't yet? For our example, this represents the risk of an immediate loan chargeoff for a loan that hasn't yet charged off after a certain amount of time.
   Cumulative hazard function: $H(t) = \int_0^t h(u) du$ represents cummulative risk up to a certain time.
   Relationship between cumulative hazard and survival: $S(t) = e^{-H(t)}$, meaning that higher hazard results in lower probability of survival.
   Data for a particular individual:
   $\delta_i$: the censoring indicator, which is 1 if individual $i$ had the event and 0 if they were censored.
   $t_i$: time of event occurrence or of censoring
   $x_i$: the covariates (predictors)
   The likelihood function: $$L = \Pi_{j \ had \ event}\ f(t_j)\ \Pi_{k \ censored}\ S(t_k) = \Pi^N_{i = 1}\ h(t_i)^{\delta_i}S(t_i)$$
   This is set up such that individuals for whom we have the time of the event (time to loan chargeoff for our purposes) are used to calculate both the hazard rate at that time and the survival to that time, but individuals who were censored count only for calculating survival to the time where they were censored.
 What happens in Cox regression:
  The Cox proportional hazards model:
  $$h(t;x) = h_0(t)\ e^{\beta x}$$
  This is for a single predictor/covariate. In the case of multiple covariates, it looks like this:
  $$h(t;x_1, ..., x_p) = h_0(t)\ e^{\beta_1x_1+...+\beta_px_p}$$
  The baseline hazard, or $h_0(t)$, is the hazard when all covariates are equal to zero. The Cox regression assumes that the ratio of the baseline hazard to the hazard function, known as the hazard ratio, is constant over time. This will be discussed in more detail in the assumptions section.
  $x_k$ represents a covariate, and $\beta_k$ is a parameter representing how the covariate affects the outcome. To interpret: As $x_k$ increases by one unit, the relative risk (hazard) is multiplied by $e^{\beta_k}$, holding all else constant. This means that covariates are having an additive effect on the log hazard ratio.
How $\beta$ values are estimated for Cox regression:
  Partial likelihood, with the following equation: $$PL(\beta) = \Pi_{t_i: event\  at\  t_i}\ \frac{h_0(t)e^{\beta x_{(t_i)}}}{\sum_{j\epsilon R_{(t_i)}}h_0(t)e^{\beta x_j}}$$
The product is taken over ordered event times. It represents, in essence, the probability that the individual for whom an event happened at a specific time were to have it happen out of all the individuals also at risk during that time. Or, in other words, the probability it happened in that case instead of any of the other potential cases.
$R_{(t_i)}$ is the risk set, all those who have neither been censored nor had the event occur at a time directly before time $t_i$.
$x_{(t_i)}$ is the value of the covariate for the individual for whom the event occurred at time $t_i$. This implicitly assumes that only one individual had the event occur at any given time, but adjustments to deal with that are implemented within the r code and will not be described in detail. 
The baseline hazard can be canceled out from the numerator and denominator of the partial likelihood, leaving this simplified equation:$$PL(\beta) = \Pi_{t_i: \ event\  at\  t_i}\ \frac{e^{\beta x_{(t_i)}}}{\sum_{j:\ t_j\geq t_i}e^{\beta x_j}}$$
We treat this partial likelihood as a likelihood, which Cox (1975) showed to be statistically supported. The standard errors for $\beta$ estimates are calculated based on a normal distribution, but this is only approximately accurate. Larger samples make the approximation more accurate, so a small sample size for a Cox regression should be treated with caution. An approximate normal distribution is also used for creating confidence intervals for our estimated hazard ratios.
The fact that we assume a particular type of relationship between our explanatory variables and survival times but don't assume anything about the shape of survival times makes this method semi-parametric.

Proportional Hazards Assumption:
  The primary assumption of a Cox regression is that the ratio between overall hazard and baseline hazard functions remains constant over time. When we create a Cox proportional hazards model, we need to check the plausibility of this assumption.
  There is a test using $\chi^2$ and the Schoenfeld residuals of our model that can be used to assess proportionality. For this test: $H_0:$ Baseline hazard is proportional to the hazard function over time $t$. $H_A:$ Baseline hazard is not proportional to the hazard function over time $t$. Notice that the null is that our assumption is met, meaning this test alone should not be relied upon to check our assumption. Higher p-values here are better, but we should also look at another method.
  Another option is to graph the scaled Schoenfeld residuals against time. If the variance is fairly consistent and no clear patterns emerge, it suggests our assumption is plausible. (We look at this similarly to residual plots we've used before, just with Schoenfeld residuals.)
  If our assumption doesn't seem to be met, there are two options. First, we could change the covariates. If we don't want to do that, we can also try stratifying by a categorical variable. That means separating individuals based on some category or strata ($k$). A different baseline hazard can be used for each stratum $k$ to account for differences between the groups, while the covariate effects are the same across all strata. This does not allow us to compare between the groups, but it allows us to interpret our covariates and apply that to all groups.
  


#Code
<!--
Here's some basic code to get the data loaded. We'll probably rename and move this section later.
-->

```{r}
#| label: Load Packages and working directory
#| code-fold: true
#| code-summary: "Show code"

suppressPackageStartupMessages(library(math3190package))
suppressPackageStartupMessages(library(survival))
suppressPackageStartupMessages(library(survminer))
suppressPackageStartupMessages(library(GGally))
suppressPackageStartupMessages(library(car))

#note, this will need to be changed to your working directory here and also at the top of the quarto document if you want to use the working directory past this chunk.

#I set up variables so we only need to change one word to get it working after pulling it. put your working directory as a string here.
kay = "~/Documents/git_connect/Predicting_Time_to_Charge_Off_Using_Cox_Regression"
samuel = "C:\\Users\\Samuel\\Desktop\\R-scripts-and-data\\MATH3190\\Final\\Cleaned_Data"
evelyn = ""

# put your name here so it takes your working directory
wd = samuel

setwd(wd)
Chargeoff_Data = read.csv("Cleaned_Chargeoff_Data.csv")
```

```{r}
#| label: convert data to numeric
head(Chargeoff_Data)
#numeric data is read as a string so we'll mutate it to make it numeric
Chargeoff_Data_Numeric = mutate(Chargeoff_Data, Effective_Annual_Interest_Rate = as.numeric(Effective_Annual_Interest_Rate), Effective_Monthly_Intrest_Rate = as.numeric(Effective_Monthly_Intrest_Rate), Time_To_Maturity = as.numeric(Time_To_Maturity), Time_to_Charge_Off = as.numeric(Time_to_Charge_Off), Original_Loan_Amount= as.numeric(Original_Loan_Amount), Monthly_Payment = as.numeric(Monthly_Payment), Total_Chargeoff = as.numeric(Total_Chargeoff), LTD_Recovery_Amount = as.numeric(LTD_Recovery_Amount), Is_Censored = as.numeric(Is_Censored))
```


```{r}
#| label: summary of data
summary(Chargeoff_Data_Numeric)
```

```{r}
#| label: check how data is distribuited

histogramMaker = function(values,x,y,title){
  #makes a bar chart for the States
  df = data.frame(values)
ggplot(df, aes(x = values)) +
  geom_histogram(color = "black", fill = "steelblue") +
  lims(x = c(min(values)-0.05, max(values)+0.05)) +
  labs(x = x, 
       y = y,
       title = title) +
  theme(axis.title = element_text(size = 16),
        axis.text = element_text(size = 12),
        plot.title = element_text(size = 18, face = "bold"))
}

barchartMaker = function(data, x_var, fill_color = "steelblue", title = NULL) {
  ggplot(data, aes(x = {{ x_var }})) +
    geom_bar(fill = fill_color) +
    labs(
      title = title,
      x = substitute(x_var),
      y = "Count"
    ) +
    theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1),
        legend.position = "None") 
}

histogramMaker(Chargeoff_Data_Numeric$Effective_Annual_Interest_Rate ,"Effective annual rate","Frequency","")

histogramMaker(Chargeoff_Data_Numeric$Time_To_Maturity ,"Time to Maturity","Frequency","")

histogramMaker(Chargeoff_Data_Numeric$Time_to_Charge_Off ,"Time_to_Charge_Off","Frequency","")

histogramMaker(Chargeoff_Data_Numeric$Monthly_Payment ,"Monthly Payment","Frequency","")

histogramMaker(Chargeoff_Data_Numeric$Original_Loan_Amount ,"Loan amount","Frequency","")

barchartMaker(Chargeoff_Data_Numeric, Is_Censored ,title = "Application codes")

barchartMaker(Chargeoff_Data_Numeric, Application_Code ,title = "Application codes")

barchartMaker(Chargeoff_Data_Numeric, Product ,title = "Types of loans")
```
<!--
_NOTE-S: Might not want to include all of those for brevity but I wrote up code on all of them.
-->

```{r}
#| label: check linearity
chargeoffNoCategorical = select(Chargeoff_Data_Numeric, Effective_Annual_Interest_Rate, Monthly_Payment, Time_To_Maturity, Original_Loan_Amount, Time_to_Charge_Off)
ggpairs(chargeoffNoCategorical)
```

A log transform is likely needed for monthly payment and original loan amount because there are many outliers for those variables. I found that taking the log of Effective_Annual_Interest_Rate actually helped the model. This surprised me because taking the log of numbers between 0 and 1 actually exaggerates their differences, but doing this seemed to help some model assumptions get met.

```{r}
#| label: check log linearity
logChargeoff = mutate(Chargeoff_Data_Numeric, Monthly_Payment = log(Monthly_Payment), Original_Loan_Amount = log(Original_Loan_Amount), Effective_Annual_Interest_Rate = log(Effective_Annual_Interest_Rate))
chargeoffNoCategorical = select(logChargeoff, Effective_Annual_Interest_Rate, Monthly_Payment, Time_To_Maturity, Original_Loan_Amount, Time_to_Charge_Off)

ggpairs(chargeoffNoCategorical)
```

There is a strong linear correlation between the log payment and log loan amount. This could cause issues with cox regression, as it is similar to linear regression.

```{r}
#| label: Cox regression model
coxMod <- coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + Time_To_Maturity + Original_Loan_Amount+ Application_Code, data = logChargeoff)
coxMod
```

A few seem to be significant in this model, but we need to check if the model assumptions are met first before we can be confident that these coefficients are reliable.

<!--
_NOTE-S: More info here: https://www.sthda.com/english/wiki/cox-model-assumptions
-->

```{r}
#| label: Testing Proportional Hazards assumption
testPH <- cox.zph(coxMod) 
testPH # If p values are high, we can assume the proportional hazards assumption is met.
```

Our p values are all Bad! This indicates that the model doesn't meet the proportional hazards assumption primarily because of the Application_Code and Time_To_Maturity variables To counteract this, we could split the data into three models by application code or stratify the model by application code.

```{r}
#| label: Stratified Cox regression model
stratCoxMod <- coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + Time_To_Maturity + Original_Loan_Amount + strata(Application_Code), data = logChargeoff)
stratCoxMod

testPH <- cox.zph(stratCoxMod) 
testPH 
```

The p values are better now that we have stratified the application codes, but the Time_To_Maturity and Original_Loan_Amount variables are causing issues with the proportional hazards assumption. We can also see this graphically by graphing the Schoenfeld residuals against the transformed time. The residuals should be randomly scattered and should form a horizontal line with no real trend as time continues. 

```{r}
#| label: Graph Cox regression model Schoenfeld residuals
ggcoxzph(testPH)
```

The graphs seem mostly good. There is a slight trend downwards with the EAR, but this shouldn't be a major concern.
Now let's check for influential observations. We do this by plotting the residuals and seeing if they are centered around zero.

```{r}
#| label: Graph Cox regression model influential obsrvations
ggcoxdiagnostics(stratCoxMod, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())

ggcoxdiagnostics(stratCoxMod, type = "deviance",
                 linear.predictions = FALSE, ggtheme = theme_bw())
```

It is fairly symmetric around zero, so we can assume this assumption is reasonably met. 
Lastly, we need to check if the variables have a linear form. We can check this by plotting the continuous covariates against martingale residuals of the cox proportional hazards model. 

```{r}
#| label: Graph Cox regression model linearity
ggcoxfunctional(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + Time_To_Maturity + Original_Loan_Amount, data = logChargeoff)
```

The variables don't look great. We will likely need to change our model to better meet these assumptions. One potential issue is VIF.

```{r}
#| label: Cox regression model VIF
vif(stratCoxMod)
```

It appears that some of our VIFs are well above 5. To compensate, we will drop Time_To_Maturity and Original_Loan_Amount, because the monthly payment is more indicative of if the borrower will be able to make payments and uses both dropped variables in its calculation. 

```{r}
#| label: Cox regression model better VIF
VifCoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + strata(Application_Code), data = logChargeoff)
VifCoxMod
vif(VifCoxMod)
```

The VIFs look far better now! Let's check the rest of the assumptions.

```{r}
#| label: Testing assumptions for Vif model
testPH = cox.zph(VifCoxMod) 
testPH # If p values are high, we can assume the proportional hazards assumption is met.

ggcoxzph(testPH)

ggcoxdiagnostics(VifCoxMod, type = "dfbeta",
                 linear.predictions = FALSE, ggtheme = theme_bw())

ggcoxdiagnostics(VifCoxMod, type = "deviance",
                 linear.predictions = FALSE, ggtheme = theme_bw())

ggcoxfunctional(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment, data = logChargeoff)
```

We can also check the linearity assumption by plotting the martingale residuals vs the payments. the relation should be mostly linear.

```{r}
#| label: Graph Cox regression model linearity Testing 

# 1. Fit the model
res.cox = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + strata(Application_Code), data = logChargeoff)
# 2. Calculate Martingale residuals
mart_res <- residuals(res.cox, type = "martingale")
martChargeoff = mutate(logChargeoff, martRes = mart_res)
# 3. Plot residuals against the predictor
ggplot(martChargeoff, aes(x = Monthly_Payment, y = mart_res)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") + # Adds a trend line
  theme_minimal() +
  labs(title = "Martingale Residuals vs Monthly_Payment",
       y = "Martingale Residuals")

# 1. Fit the model
res.cox <- coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + strata(Application_Code), data = logChargeoff)
# 2. Calculate Martingale residuals
mart_res <- residuals(res.cox, type = "martingale")
martChargeoff = mutate(logChargeoff, martRes = mart_res)
# 3. Plot residuals against the predictor
ggplot(martChargeoff, aes(x = Effective_Annual_Interest_Rate, y = mart_res)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess") + # Adds a trend line
  theme_minimal() +
  labs(title = "Martingale Residuals vs Effective_Annual_Interest_Rate",
       y = "Martingale Residuals")


```

Most of the assumptions look good, but the linearity assumption doesn't look perfect. It's about as close as I can get it to being linear, so we'll go with it despite that.

Now lets try making separate models for each application code to see if they systematically differ in the way interest rate and loan payment affect them.

```{r}
#| label: Cox regression by Application_Code 
CCdata = filter(logChargeoff, Application_Code == "CC")
CRdata = filter(logChargeoff, Application_Code == "CR")
CIdata = filter(logChargeoff, Application_Code == "CI")

CC_CoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment, data = CCdata)
CC_CoxMod

CR_CoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment, data = CRdata)
CR_CoxMod

CI_CoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment, data = CIdata)
CI_CoxMod
```

It seems that the coefficients of the models are mostly identical, with one exception being that the effect of Monthly_Payment is a little smaller in Consumer Installment loans. Using stratification makes sense here because individually, we don't get as much data as if we stratify the model and it would appear that the different application types don't systematically differ in the way the variables affect them.  

Let's look at the Survival curve for our final stratified model:

```{r}
#| label: survival curve 
FinalCoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + strata(Application_Code), data = logChargeoff)

fit <- survfit(FinalCoxMod, data = logChargeoff)
ggsurvplot(fit, 
           risk.table = TRUE,
           surv.median.line = "hv", 
           data = logChargeoff, 
           conf.int = TRUE,         
           palette = "Set1",          
           legend.title = "Application Code", )

```


A couple things to note about the survival curve, 
1. there were four loans that had a time to charge off of more than 5500 days (about 15 years), one of which was a $15.90, 2 year loan from 1982 that lasted over 11,000 days before being charged off. I can't imagine there was any hope this person was tying to pay back their loan 30 years after the loan was issued, so we could use an administrative censor to censor these points at 6000 days after their issuing. This is generally better for our model since we won't have these extreme outliers, and it will make our graph far easier to observe

```{r}
#| label: censor extreme outliers and redo survival curve
logChargeoffClean = mutate(logChargeoff, Is_Censored = ifelse(Time_to_Charge_Off > 5500, 1, Is_Censored), Time_to_Charge_Off = ifelse(Time_to_Charge_Off > 5500, 5500, Time_to_Charge_Off))

FinalCoxMod = coxph(Surv(Time_to_Charge_Off, Is_Censored) ~ Effective_Annual_Interest_Rate + Monthly_Payment + strata(Application_Code), data = logChargeoffClean)

fit <- survfit(FinalCoxMod, data = logChargeoff)
ggsurvplot(fit, 
           risk.table = TRUE,
           surv.median.line = "hv", 
           data = logChargeoff, 
           conf.int = TRUE,         
           palette = "Set1",          
           legend.title = "Application Code", )
```


